Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-02-01 ms.tex
edited appendix D

% Remus commands
\newcommand{\x}{{\bf x}}
\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\newcommand{\erfc}{{\rm erfc}}
\newcommand{\ii}{{\rm i}}

@@

\section{Recurrence time}\label{sec:appendix_recurrence_time} 
In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll 1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$). 
We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}),
\begin{equation*}
n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*} .
\end{equation*}
We use \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\; \abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.

@@

In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of oridnary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}
This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that $\e^{\Delta_m \tau_a^r} \gg\e^{\Delta_s \tau_a^r},\e^{\Delta_a \tau_a^r}$ for even small values of $\tau_a^r$ because $\Delta_m>0$ and $\Delta_s,\Delta_a<0$. 
We assume that the tumor is large ($N\gg N_a^*$) % REMUS you assumed N\gg N_m^* not N\gg N_a^*
so the time $\tau_a^r$ for the resistant cell population to reach $N$ is large enough such that the first term of \cref{eq:t2det2024} can be written as
\begin{align*}
\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)} ,
\end{align*}
which is significantly smaller then the second term of \cref{eq:t2det2024},
\begin{align*}
v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)} .
\end{align*}
Consequently, \cref{eq:t2det2024} can be approximated by
\begin{equation*}
1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\Delta_m - \Delta_s},
\end{equation*}
which has an approximate solution
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We observe that the terms given by the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N \gg N_m^* > N_a^*$. %change 9

@@

\section{Distribution of evolutionary rescue time}\label{sec:appendix_distribution_time}
The probability that a successful mutant has been generated by time $t$ is given by:
\begin{align*}
P\left(rescue,t\right)&=P\left(T_a<t\right)\\
&=1-\exp\left\{-\left[n_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
&=1-\exp\left\{-\left[\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\right\},
\end{align*}
where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in appendix \ref{sec:appendix_rescue_time}.

>>>>

% Remus commands
\newcommand{\x}{{\bf x}}
\renewcommand{\d}{{\rm d}}
\newcommand{\e}{e}
\newcommand{\erfc}{{\rm erfc}}
\newcommand{\ii}{{\rm i}}

@@

\section{Recurrence time}\label{sec:appendix_recurrence_time}

In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}),
\begin{equation*}
n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*} .
\end{equation*}
We use \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\; \abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.

@@

In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

\eqref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the final tumor size $N$ is very large, and that since the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells. 
Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
\eqref{eq:t2det2024} can then be rewritten as:
\[
	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
\]
Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants (i.e., that $u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding:
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation. %change 9

@@

\section{Distribution of evolutionary rescue time}\label{sec:appendix_distribution_time}

The probability that a successful mutant has been generated by time $t$ is given by:
\begin{align*}
P\left(rescue,t\right)&=P\left(T_a<t\right)\\
&=1-\exp\left\{-\left[n_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
&=1-\exp\left\{-\left[\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\right\},
\end{align*}
where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in appendix \ref{sec:appendix_rescue_time}.
